A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-07 podc_paper/algorithm-2.tex
minor edits; changed tournament tree to ordering tree
working on GC

Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
We first describe the modifications to the implementation.
It uses two additional shared arrays.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ indexes last block in $root$ that process $k$ has dequeued from}
\end{itemize}
The \fld{blocks} field is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}},\\
		\gc{\fld{index} \assign\ \var{leaf}.\head}}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1,\\
	    \gc{\fld{index}\assign \var{leaf}.\head}}
    \State \Call{Append}{\var{B}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}
    \label{deqRestGC}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \State \gc{\var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHeadGC}

\gc{    
	\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
	\State \var{max} \assign 0
	\For {$k \assign 1..p$}
		\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
	\EndFor 
	\State \Call{Help()}{}
	\State \Call{Discard}{$\var{root}, \var{max}-1$}
	\EndIf
	\State
}

    \ForEach{\fld{dir} {\keywordfont{in}} \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHeadGC}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNullGC}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvanceGC}
        \EndIf
    \EndFor \label{endHelpChild1GC}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC} 
    \EndIf
    \State \gc{\var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
    \State \var{\Call{Advance}{\var{v, h}}}\label{advanceGC}
    \State \Return{ \var{result}}
\EndFunction{Refresh}

@@

\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} 
    \State \linecomment{set \var{v}.\fld{blocks}[\var{h}].\fld{super} and }
    \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
    \If{$\var{v}\neq \var{root}$}
        \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHeadGC}
        \State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1GC}
    \EndIf
    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHeadGC}
\EndFunction{Advance}

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \gc{\var{new}.\fld{index} \assign \var{h}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
        \State \gc{\var{last[this process id]} \assign\ \var{b}}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
        \State \gc{\var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}}\label{findAnswerGC}
        \State \gc{\var{last[this process id]} \assign\ $b_e$}
        \State \gc{\Return{\var{e}}}
    \EndIf
\EndFunction{FindResponse}

\end{algorithmic}
\end{minipage}
\caption{Queue implementation.\label{pseudocode1GC}}
\end{figure}

@@

\gc{
\Function{void}{Help}{}
\State \bf{for} \var{i} \assign\ 1 to $p$
\State h=leafOfProcess[\var{i}]
\If{leafOfProcess[\var{i}].blocks[h].num\sub{deq}==1 and leafOfProcess[\var{i}].IndexDequeue(h,1)!=null}
\State <b, i> \assign\ \Call{IndexDequeue}{leafOfProcess[\var{i}], h, 1}
\State output \assign\ \Call{FindResponse}{b, i} 
\State leafOfProcess[\var{i}].blocks[h].response \assign\ output
\EndIf
\State \bf{end for}
\EndFunction{Help}

\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{i}}
    \State $T \assign \var{v}.\fld{blocks}$
    \State $\langle T_1, T_2\rangle \assign \Call{Split}{T,i}$
    \State \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T_2$}
    \If{\var{v} is an internal \typ{Node}}
        \State \Call{Discard}{\var{v}.\fld{left}, \var{v}.\fld{blocks}[\var{i}].\eleft}
        \State \Call{Discard}{\var{v}.\fld{right}, \var{v}.\fld{blocks}[\var{i}].\eright}
    \EndIf
\EndFunction{Discard}
}
\end{algorithmic}
\caption{\label{pseudocode2GC}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
\end{figure}

If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.

>>>>

Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
We first describe the modifications to the implementation.
It uses two additional shared arrays.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ indexes last block in $root$ that process $k$ has dequeued from}
\end{itemize}
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}
The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  Each RBT is initialized with an empty block with index 0.

\here{Talk about what happens if a block is not found in an RBT}

We also apply an optimization:  instead of storing a \head\ field separate from \fld{blocks},
we can simply use the maximum \fld{index} of any block in the \fld{blocks} RBT.

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
	\gc{\State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}}+1,$\\
		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}},$\\
		$\fld{index} \assign\ h+1$}}\label{enqNewGC}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
	\gc{\State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}},$\\
	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}+1,$\\
	    $\fld{index}\assign h+1$}}\label{deqNewGC}
    \State \Call{Append}{\var{B}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \gc{$h+1$}, 1}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}
    \label{deqRestGC}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{Insert}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
    \gc{\State \here{Need to clean out my own leaf sometimes--how to determine what threshold to use for the split?}}
\EndFunction{Append}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{\gc{GCRefresh}}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
        \State \Call{\gc{GCRefresh}}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

	\gc{
\Function{boolean}{GCRefresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}, with GC}
	\State $\var{T} \assign\ \var{v}.\fld{blocks}$\label{readTGC}
    \State $\var{h} \assign\ \Call{MaxBlock}{T}.\fld{index}$ \label{readHeadGC}
	\If{\var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
		\State \linecomment{garbage collection phase}
		\State \var{max} \assign 0
		\For {$k \assign 1..p$}
			\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
		\EndFor
		\State \Call{Help}{}
		\State \Return $\Call{Discard}{\var{root}, \var{max}-1}$\label{discard}
    \Else\ \linecomment{do a standard refresh}
        \State $T' \assign \Call{RefreshUpdate}{T, v}$\label{RU1}
        \If{$T' = \nl$} \Return{\tr}
    	\Else\ \Return $\Call{CAS}{\var{v}.\fld{blocks}, T, T'}$\label{casGC}
		\EndIf
	\EndIf
%    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
\EndFunction{GCRefresh}
}

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}
\gc{\Function{RBT}{RefreshUpdate}{\typ{RBT} \var{T}, \typ{Node} \var{v}}
    \State \linecomment{append new block to RBT $T$ read from $v.\fld{blocks}$}
    \State \Call{SetSuper}{$v$, $T$}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
    \Else \ \Return \Call{Insert}{$T$, \var{new}}\label{refreshInsertGC}
    \EndIf
\EndFunction{RefreshUpdate}
}
\spac

\gc{
\Function{void}{SetSuper}{\typ{Node} \var{v}, \typ{RBT} $T$} 
	\State \linecomment{set \fld{super} field of last block in $T$}
	\State \linecomment{and in $v$'s children's RBTs}
	\State $\var{lastBlock} \assign \Call{MaxBlock}{T}$
	\If{$v\neq \var{root}$}	
		\State \var{sup} \assign\ \Call{MaxBlock}{\var{v}.\fld{parent.blocks}}.\fld{index}
		\State \Call{CAS}{\var{lastBlock}.\fld{super}, \nl, \var{sup}}
	\EndIf
	\ForEach{\fld{dir} in \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
    	\State \var{lastChildBlock} \assign\ \Call{MaxBlock}{\var{v}.\fld{dir}.\fld{blocks}}
		\State \Call{CAS}{\var{lastChildBlock}.\fld{super}, \nl, \var{lastBlock}.\fld{index}}
    \EndFor \label{endHelpChild1GC}
\EndFunction{SetChildrenSuper}
}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \gc{
    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
    \State \var{new}.\fld{index} \assign \var{i}
    }
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}
\end{algorithmic}
\end{minipage}
\caption{Bounded-space queue implementation.\label{pseudocode1GC}}
\end{figure}

\begin{figure}
\begin{minipage}[t]{0.465\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{48}
\gc{
\Function{void}{Help}{}
    \State \linecomment{help pending operations}

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{last} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
    	\State \Call{Propagate}{$\ell.\fld{parent}$}
        \If{$\var{last}.\fld{element}\hspace{-.1mm} =\hspace{-.1mm} \nl$\hspace{-.15mm} and\hspace{-.15mm} $\var{last}.\fld{index}\hspace{-.12mm}>\hspace{-.12mm}0$} 
            \State \linecomment{operation is a  dequeue}
            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, \var{last}.\fld{index}, 1}$
            \State $\var{last}.\fld{response} \assign\ \Call{FindResponse}{b, i}$ 
        \EndIf
    \EndFor
\EndFunction{Help}

@@

\Function{boolean}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
    \State \linecomment{discard blocks with index less than or equal}
    \State \linecomment{to $t$ from $v$'s RBT and their subblocks}
    \If{$v$ is an internal node}
	    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
    	\State \var{t\sub{right}} \assign \var{v}.\fld{blocks}[\var{t}].\eright
		\For{\var{j} \assign 1 to $p$}
        	\State $T \assign \var{v}.\fld{blocks}$\label{readT2GC}
        	\State $T'  \assign \Call{Split}{T,t}$\label{split}
			\State \linecomment{\op{Split} removes blocks with $\var{index} < t$}
			\State $T'' \assign \Call{RefreshUpdate}{T',v}$ \label{RU2}
        	\State \hangbox{$\var{result} \assign \Call{CAS}{\var{v}.\fld{blocks}, T, \\
					\hspace*{10mm}(T'' = \nl\ ?\ T' : T'')$}}\label{splitCASGC}
        	\State \bf{exit when} \var{result} 
	        \State \linecomment{if \op{CAS} succeeds go to line \ref{discardrecursecomment}}
    	\EndFor
    	\If{\var{v} is an internal \typ{Node}} 
			\State \linecomment{discard subblocks in descendants of $v$}\label{discardrecursecomment}
        	\State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}\label{discardrecurse1}
        	\State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}\label{discardrecurse2}
    	\EndIf
	\EndIf
	\State \Return \var{result}
\EndFunction{Discard}
}
\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
        \gc{\State \var{last}[\var{this process id}] \assign\ \var{b}}\label{setlast1}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
        \gc{
        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
        \State \var{last}[\var{this process id}] \assign\ $b_e$\label{setlast2}
        \State \Return{\var{e}}
        }
    \EndIf
\EndFunction{FindResponse}

\end{algorithmic}
\end{minipage}

\caption{Bounded-space queue implementation, continued.\label{pseudocode1GC}}
\end{figure}

@@

\end{algorithmic}
\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
\end{figure}

\subsection{Correctness}
Unfortunately, there are enough changes to the algorithm that a new proof of correctness
is required.  Its structure mirrors the proof of the original algorithm, but requires additional
reasoning to ensure GC does not interfere with other routines.

\begin{observation}\label{RBTinsertions}
When a block is inserted into an RBT $T$, its \fld{index} is one greater than the maximum \fld{index} in $T$.
\end{observation}
\begin{proof}
If line \ref{appendLeafGC} inserts a block into a leaf's RBT, then it was created at 
line \ref{enqNewGC} or \ref{deqNewGC} with \fld{index} $h+1$, where $h$ is the maximum \fld{index}
in the leaf's RBT.  When the new block is inserted, the maximum \fld{index} in the RBT is still $h$,
since only the owner of the leaf modifies the leaf's RBT.

If line \ref{refreshInsertGC} inserts a block  into an internal node's RBT $T$,
it is created by the call to \op{CreateBlock} at line \ref{invokeCreateBlockGC} and 
has \fld{index} $h$, which is set at line \ref{refreshReadMax} to  the maximum \fld{index} in $T$ plus one.
\end{proof}

\begin{observation}\label{lastinroot}
Each entry in \var{last} is the \var{index} of a block that has appeared in the root's RBT.
\end{observation}
\begin{proof}
Only line \ref{setlast1} and \ref{setlast2} in \op{FindResponse} write values into \var{last}.
The value written is either $b$ or $b_e \leq b$.
By the precondition of \op{FindResponse}, $\var{root}.\fld{blocks}[b]$ exists, so the index $b$ has appeared
in the root's RBT.
By \Cref{RBTinsertions}, all values less than $b$ have appeared in the root's RBT too.
\end{proof}

\begin{observation}
\label{edirGC} 
If $B$ is a block in the RBT of an internal node $v$, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
$B.\edir$ was read from the \fld{index} field of some node in the RBT of $v.\var{dir}$.
\end{observation}
\begin{proof}
The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
\end{proof}

\begin{invariant}
For each RBT $T$, 
$T$ is non-empty and the blocks of $T$ have consecutive \fld{index} fields.
\end{invariant}
\begin{proof}
The claim is true initially, since each RBT initially contains only the empty block with \fld{index} 0.
We show that each time a new RBT is constructed, it preserves the invariant.
When a new RBT is created by inserting an element into an existing RBT, \Cref{RBTinsertions}
implies that the invariant is preserved.

Suppose a new RBT is created by a \op{Split}($T,t$) operation $S$ on at line \ref{split} of a \op{Discard}($v,t$), 
and that the invariant held prior to $S$.
Clearly, $S$ preserves the property that indices are consecutive.  To show that the new tree
is non-empty we first prove that the call to \op{Discard}($v,t$) has a parameter $t$ that is at least as big as a value previously
read from the index field of some block in $v$'s RBT.
If \op{Discard} was called from line \ref{discard}, then $t+1$ was read from an entry of \var{last}, so the claim follows from \Cref{lastinroot}.
If \op{Discard} was called from line \ref{discardrecurse1} or \ref{discardrecurse2}, the claim follows
from \Cref{edirGC}.
Since the invariant held prior to $S$, no split before $S$ emptied $v$'s tree.  Therefore,
some block with index at least $t$ is still in $T$ when $S$ occurs.  That block will be in the new tree created by $S$.
\end{proof}

Thus, calls to \op{MaxBlock} are well-defined.  Also, this means that the sequence of blocks added to $v.\fld{blocks}$ have consecutive \fld{index} values, starting from 1, so the references to $v.\fld{blocks}[i]$ (meaning the block in $v$'s RBT with \fld{index} $i$) are well-defined.

\begin{corollary}\label{maxNonDecreasing}
The maximum \fld{index} in $v$'s RBT is non-decreasing over time.
\end{corollary}
\begin{proof}
The CAS steps (on line \ref{casGC} and \ref{splitCASGC}) that update $v.\fld{blocks}$ always replace 
a tree $T$ by a tree that is built from $T$ by inserting an element or removing the elements with 
the smallest \fld{index} values.
\end{proof}

We now prove the analogue of Lemma \ref{lem::headProgress}.
\begin{lemma}
 If $b>0$ and a block with index $b$ has been inserted into $v.\fld{blocks}$ then 
 $v.\fld{blocks}[b-1].\fld{end\sub{left}} \leq v.\fld{blocks}[b].\fld{end\sub{left}}$ and 
 $v.\fld{blocks}[b-1].\fld{end\sub{right}} \leq v.\fld{blocks}[b].\fld{end\sub{right}}$.
\end{lemma}
\begin{proof}
The block $B$ with index $b$ was installed into $v$'s RBT by the CAS at line \ref{casGC} or \ref{splitCASGC} 
that changed the tree from $T$ to $T_{new}$.
Before this CAS, the tree $T$ was read from $v.blocks$ at line \ref{readTGC} or \ref{readT2GC}, respectively,
before the \op{RefreshUpdate} on line \ref{RU1} or \ref{RU2} that inserted $B$ to get $T_{new}$.
Since the previous block $B'=v.\fld{blocks}[b-1]$ was already in $T$ (by \Cref{RBTinsertions}),
the \op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the \op{CreateBlock}$(v,b)$ that created $B$ started.
It follows from \Cref{maxNonDecreasing} that the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.  Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
\end{proof}

\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} hold true for the modified algorithm.  Their proofs are identical to those given in Section \ref{sec::basicProperties}.

-------
